---
title: "Model report"
author: "Jessica Guo"
date: "6/15/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(udunits2)
library(dplyr)
library(coda)
library(broom.mixed)
library(ggplot2)
```


#### Introduction
The temporal repackaging of rainfall experiment at the Santa Rita Experimental Range ('RainManSR') utilizes a random block design to test the influence of rainfall intensity and frequency on soil, plant, and ecosystem dyanamics, while holding total rainfall amount constant. The soil water content (SWC) of most plots are monitored at three depths 0-12 cm, 25 cm, and 75 cm. In two individual plots representing the S1 (frequent small rain events) and the S4 (infrequent large rain events), soil water potential (SWP) probes were additionally installed, allowing for us to generate <i>in-situ</i> soil moisture release curves. 


#### Data description
All soil moisture data can be interactively explored via the [Shiny app](https://viz.datascience.arizona.edu/rainman-soildata/). For this analysis, we selected the 25 cm depth, for which A) a range of moisture values was achieved and B) the SWC and SWP sensors were installed at identical depths. For both the S1 and S4 treatments, we used the late winter and early spring data (2019-12-01 to 2020-04-30) for the 25 cm depth (n = 152 for each S1 and S4). 


#### Model description

We used the Van Genuchten function to predict SWC from SWP. The observed SWC($\Theta_i$) were normally distributed ($i = 1 \dots 304$):

$\Theta_i \sim Normal(\mu_i, \tau)$

The expected value was modeled non-linearly as:

$\mu_i = \Theta_{r, trt(i)} + \frac{\Theta_s - \Theta_{r, trt(i)}}{[1 + (\alpha_{trt(i)}\Psi_i)^{n_{trt(i)}}]^{1-\frac{1}{n_{trt(i)}}}}$

Where $\Theta_s$ was a constant 0.44 (porosity of loamy sand), and the parameters $\Theta_r$, $\alpha$, and $n$ were indexed by irrigation treatment $trt(i)$ and modeled hierarchically.

Treatment-level parameters were modeled with lognormal distributions centered at the site-level parameter. 

$\Theta_{r, t} \sim dlnorm(\mu_{log(\theta_r)}, \tau_{log(\theta_r)})$

$\alpha_t \sim dlnorm(\mu_{log(\alpha)}, \tau_{log(\alpha)})$

$n_t \sim dlnorm(\mu_{log(n)}, \tau_{log(n)})$

For the site-level parameters on the log scale, semi-informative but wide normal priors were specified:

$\mu_{log(\theta_r)} \sim dnorm(-3.5, 3.2)$ 

$\mu_{log(\alpha)} \sim dnorm(-2.5, 1.5)$

$\mu_{log(n)} \sim dnorm(0.65, 3)$

The above equations were programmed in JAGS; three parallel chain were run for 150,000 iterations with thinning by 150 to reduce storage space for a total of 3000 samples. Parameters are summarized as posterior means and the central 95% credible interval (CI). If the posterior mean of one parameter is not overlapped by the CI of the other and vice versa, then the two parameters are said to be significantly different. 

```{r, echo=FALSE}
# Load data 
load("out.Rdata")

dat <- filter(out, Depth == 2) %>%
  mutate(pressure_head_cm = ud.convert(WP_mean, "MPa", "cm_H2O"),
         pressure_head_m = ud.convert(pressure_head_cm, "cm", "m"))

# Load codas
load(file = "coda/jm_coda.Rdata")
load(file = "coda/jm_rep.Rdata")

# Tidy parameters
sum_param <- tidyMCMC(as.mcmc(jm_coda[[1]], jm_coda[[2]], jm_coda[[3]]),
                      conf.int = TRUE,
                      conf.method = "HPDinterval")

# # Tidy replicated and add to original data
sum_rep <- tidyMCMC(as.mcmc(jm_rep[[1]], jm_rep[[2]], jm_rep[[3]]),
                    conf.int = TRUE,
                    conf.method = "HPDinterval")

pred <- cbind.data.frame(dat, sum_rep)
```


#### Model fit

The observed vs. predicted indicate good model fit for overall, R<sup>2</sup> = 0.98. The dashed line represents the 1:1 line. 

```{r, echo=FALSE, fig.height=4, fig.width=4}
m <- summary(lm(WC_mean ~ estimate, data = pred))

# Observed vs. fitted
pred %>%
  ggplot(aes(x = WC_mean, y = estimate)) +
  geom_abline(slope = 1, intercept = 0, lty = 2) +
  geom_errorbar(aes(ymin = conf.low, 
                    ymax = conf.high,
                    col = Summer),
                width = 0,
                alpha = 0.25) +
  geom_point(aes(col = Summer), size = 2) +
  scale_color_manual(values = c("goldenrod", "forestgreen")) +
  scale_x_continuous("Observed SWC") +
  scale_y_continuous("Predicted SWC") +
  # facet_wrap(~Summer) +
  theme_bw(base_size = 14) +
  theme(legend.title = element_blank(),
        legend.position = c(0.85, 0.2),
        panel.grid = element_blank())

```
 
Model fit also looks promising when plotted along the original Van Genuchten axes of water content vs. water potential. 

```{r, echo=FALSE, fig.height=4, fig.width=8}
pred %>% 
  ggplot(aes(x = -pressure_head_m)) +
  geom_point(aes(y = WC_mean, 
                 color = "Observed")) +
  geom_pointrange(aes(y = estimate, 
                      ymin = conf.low,
                      ymax = conf.high,
                      color = "Predicted")) +
  facet_wrap(~Summer) +
  scale_color_manual(values = c("black", "gray")) +
  scale_x_log10(name = "Pressure Head (-m)",
              breaks = c(0.01, 0.1, 1, 10, 100, 1000),
              labels = c(0.01, 0.1, 1, 10, 100, 1000)) +
  scale_y_continuous("SWC") +
  theme_bw(base_size = 14) +
  theme(legend.title = element_blank(),
        legend.position = c(0.85, 0.85),
        panel.grid = element_blank())
```

Parameters $\Theta_r$, $\alpha$, and $n$ were estimated by the model at the treatment level (colors) nested within site (black). Despite the borrowing of strength among treatments at the same site, all parameters were significantly different between the two treatments, S1 and S4. 

```{r, echo=FALSE, fig.height=3, fig.width=8}
pop <- sum_param %>%
  filter(grepl("E\\.", term)) %>%
  mutate(parameter = sub("\\[[0-9]\\]", "", term),
         parameter = sub("E\\.", "", parameter),
         label = case_when(parameter == "alpha.cm" ~ "alpha",
                           parameter == "n" ~ "n",
                           parameter == "theta.r" ~ "theta[r]"),
         label = factor(label, levels = c("theta[r]", "alpha", "n")))

trt <- sum_param %>%
  filter(grepl("^alpha\\.cm", term) |
           grepl("^n\\[", term) |
           grepl("^theta.r" , term)) %>%
  mutate(parameter = sub("\\[[0-9]\\]", "", term),
         Treatment = case_when(grepl("1", term) ~ "S1",
                               grepl("2", term) ~ "S4"),
         label = case_when(parameter == "alpha.cm" ~ "alpha",
                           parameter == "n" ~ "n",
                           parameter == "theta.r" ~ "theta[r]"),
         label = factor(label, levels = c("theta[r]", "alpha", "n")))

ggplot() +
  geom_pointrange(data = trt,
                  aes(x = label,
                      y = estimate,
                      ymin = conf.low, 
                      ymax = conf.high,
                      color = Treatment),
                  position = position_dodge(width = 1)) +
  geom_pointrange(data = pop,
                  aes(x = label,
                      y = estimate,
                      ymin = conf.low, 
                      ymax = conf.high)) +
  scale_color_manual(values = c("goldenrod", "forestgreen")) +
  facet_wrap(~label, scales = "free",
             labeller = label_parsed) +
  theme_bw(base_size = 14) +
  theme(panel.grid = element_blank(),
        legend.title = element_blank(),
        axis.text.x = element_blank())
```

#### Predicting SWP from SWC

One goal of this project was to predict SWP for plots where only SWC was measured empirically. Here, we see that the posterior mean of site-level parameters appear to fit data from both treatments, although only the dry end of the relationship can be validated. 

```{r, echo=FALSE, fig.height=4, fig.width=6}
param_site <- sum_param %>%
  filter(grepl("^E\\.", term))


VG <- function(h.cm, vec){
  alpha <- as.numeric(vec[1])
  n <- as.numeric(vec[2])
  theta.r <- as.numeric(vec[3])

  wc <- theta.r + ((0.44 - theta.r)/
                            (1 + (alpha*h.cm)^n)^(1-(1/n)))
  
  return(wc)
}

out_site <- data.frame(ph_cm = seq(1, 10000, by = 1)) %>%
  mutate(WC = VG(ph_cm, vec = param_site$estimate),
         ph_MPa = ud.convert(ph_cm, "cm_H2O", "MPa"))

ggplot() +
  geom_line(data = out_site,
            aes(x = ph_cm, y = WC),
            lwd = 1) +
    geom_point(data = dat, 
             aes(x = -pressure_head_cm,
                 y = WC_mean,
                 color = Summer)) +
  scale_x_log10(name = "Pressure Head (-cm)",
                breaks = c(0.01, 0.1, 1, 10, 100, 1000, 10000),
                labels = c(0.01, 0.1, 1, 10, 100, 1000, 10000)) +
  scale_y_continuous("SWC") +
  scale_color_manual(values = c("goldenrod", "forestgreen")) +
  theme_bw(base_size = 14) +
  theme(legend.title = element_blank(),
        legend.position = c(0.9, 0.8),
        panel.grid = element_blank())

ggplot() +
  geom_line(data = out_site,
            aes(x = ph_MPa, y = WC),
            lwd = 1) +
    geom_point(data = dat, 
             aes(x = -WP_mean,
                 y = WC_mean,
                 color = Summer)) +
  scale_x_log10(name = "Pressure Head (-MPa)",
                breaks = c(0.01, 0.1, 1, 10, 100, 1000, 10000),
                labels = c(0.01, 0.1, 1, 10, 100, 1000, 10000)) +
  scale_y_continuous("SWC") +
  scale_color_manual(values = c("goldenrod", "forestgreen")) +
  theme_bw(base_size = 14) +
  theme(legend.title = element_blank(),
        legend.position = c(0.9, 0.8),
        panel.grid = element_blank())

```

We used the posterior parameters sets (n = 3000) at the site- and treatment-levels to create a lookup table and associated function `swc2swp()`to convert SWC (proportion) to SWP (MPa). The `param` argument can be set to "site" (default), "S1", or "S4", and the `stat` argument can be set to "median" (default), "lower", or "upper", where the latter options indicate the bounds of the central 50th percentile. 

Below, we demonstrate how to load the function (which includes the lookup table) and use it for a single or vector of input SWC values.  

```{r}
source("../../source/swc2swp.R")

swc2swp(SWC = 0.05, param = "site", stat = "median")
swc2swp(SWC = seq(0.05, 0.15, by = 0.01), param = "site", stat = "median")

```

By applying the function to the empirical dataset, we can compare the predictive interval of SWP to the observed SWP for site-level parameters. 

```{r, echo=FALSE, warning=FALSE, fig.height=4, fig.width=8}
dat$test_WP <- swc2swp(SWC = dat$WC_mean, stat = "median")
dat$test_WP_25 <- swc2swp(SWC = dat$WC_mean, stat = "lower")
dat$test_WP_75 <- swc2swp(SWC = dat$WC_mean, stat = "upper")

dat %>%
  ggplot(aes(y = WC_mean)) +
  geom_point(aes(x = abs(test_WP),
                 col = "predicted")) +
  geom_errorbarh(aes(xmin = abs(test_WP_25),
                     xmax = abs(test_WP_75),
                     col = "predicted"),
                 alpha = 0.25) +
  geom_point(aes(x = abs(WP_mean),
                 col = "observed")) +
  scale_y_continuous("SWC") +
  scale_x_log10(name = "SWP (-MPa)") +
  scale_color_manual(values = c("black", "gray70")) +
  facet_wrap(~Summer) +
  theme_bw(base_size = 14) +
  theme(legend.title = element_blank(),
        legend.position = c(0.85, 0.85),
        panel.grid = element_blank())

```

The site-level parameters don't fit the empirical data from either treatment particularly well. Let's see how treatment-level parameters perform. 

```{r, echo=FALSE, warning=FALSE, fig.height=4, fig.width=8}
for(i in 1:nrow(dat)) {
  dat$test_WP[i] <- swc2swp(SWC = dat$WC_mean[i], param = dat$Summer[i], stat = "median")
  dat$test_WP_25[i] <- swc2swp(SWC = dat$WC_mean[i], param = dat$Summer[i], stat = "lower")
  dat$test_WP_75[i] <- swc2swp(SWC = dat$WC_mean[i], param = dat$Summer[i], stat = "upper")
}

dat %>%
  ggplot(aes(y = WC_mean)) +
  geom_point(aes(x = abs(test_WP),
                 col = "predicted")) +
  geom_errorbarh(aes(xmin = abs(test_WP_25),
                     xmax = abs(test_WP_75),
                     col = "predicted"),
                 alpha = 0.5) +
  geom_point(aes(x = abs(WP_mean),
                 col = "observed")) +
  scale_y_continuous("SWC") +
  scale_x_log10(name = "SWP (-MPa)") +
  scale_color_manual(values = c("black", "gray70")) +
  facet_wrap(~Summer) +
  theme_bw(base_size = 14) +
  theme(legend.title = element_blank(),
        legend.position = c(0.85, 0.85),
        panel.grid = element_blank())

```

Predictions made with treatment-level parameters match the observed SWP better than those made with site-level parameters, at both the upper and lower end of the observed SWC range. However, it appears that the standard Van Genuchten model simply cannot not capture behavior at the low water contents that are typical of the RainMan experiment. 

#### Future work

While the primary goal of this work was to relate SWC and SWP to derive site-specific Van Genuchten parameters to be used in physical models, we also created the `swc2swp()` function and lookup table to predict SWP in the RainMan plots. Future work could involve any combination of the following:

1) Field experiment of paired SWC and SWP senors with heavy irrigation, to obtain a wider range of data for parameter estimation

2) Soil samples sent to lab for lab-derived relationship and comparison to this field-based approach

3) Fitting a modified Van Genuchten [model](https://www.pnnl.gov/main/publications/external/technical_reports/PNNL-19800.pdf) that extends to low saturation soils where SWC may be less than residual water content ($\theta_r$). This approach involves predicting the changepoint in addition to the non-linear Van Genuchten parameters and may be difficult to fit, especially hierarchically

4) Alternatively, to obtain the best empirical fit, a spline approach will produce high predictive performance. In that case, the `swc2swp()` function can be updated with a `method` argument and the lookup table can include predictions from the spline model
