---
title: "Initial data exploration"
author: "Jessica Guo"
date: "3/31/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Soil data from RainMan SR

This project uses soil volumetric water content (VWC) and temperature (Ts) data from the RainManSR project to predict soil water potential (SWP), which is available from a subset of plots/treatments. The project goals are two-fold: 1) to extract Van Genuchten model parameters for use in predicting SWP mechanistically, using model such as Hydrus 1D, and 2) to develop a statistical model of SWP using VWC in a hierarchical Bayesian framework and predict the missing values of SWP for other plots/treatments. 

```{r}
library(xts)
library(dplyr)
library(ggplot2)
library(RColorBrewer)
library(plotly)
library(cowplot)
library(xts)
library(dygraphs)
library(plantecophys)
```

## Data exploration

Soil temperature and soil water data are recorded as half-hourly time series in two .csv files. Depths 1 - 3 correspond with 0-12 cm, 25 cm, and 75 cm respectively. Soil water data includes VWC, SWP, and air temperature & RH both inside and outside the houses. Data are read in and and converted to long format for plotting and summarizing purposes. Three dataframes are produced: Ts (soil temperature only), WC (VWC and SWP), and Ta (air temperature, RH, and VPD). 
Air temperature below -6 or so is not reasonable, so filter out T values < -10. 

```{r}
Ts <- read.csv("../data/Halfhourly_TS_0113.csv") %>%
  dplyr::select(-X) %>%
  mutate(TIMESTAMP = as.POSIXct(TIMESTAMP),
         date = as.Date(TIMESTAMP)) %>%
  rename(dt = TIMESTAMP) %>%
  tidyr::pivot_longer(!c(dt, date), names_to = c("Plot", "Depth", "House"), 
                      names_pattern = "TS_P(.*)_(.*)_H(.*)") %>%
  mutate(House = factor(as.numeric(House)),
         Depth = factor(as.numeric(Depth)),
         Plot = factor(as.numeric(Plot)),
         ID = paste0("P", Plot, "_", Depth, "_H", House)) %>%
  arrange(ID, dt)

WC <- read.csv("../data/Halfhourly_VWC_Weather_030621.csv") %>%
  dplyr::select(-X, -T_inside, -RH_inside, -T_outside, -RH_outside) %>%
  mutate(TIMESTAMP = as.POSIXct(TIMESTAMP),
         date = as.Date(TIMESTAMP)) %>%
  rename(dt = TIMESTAMP) %>%
  tidyr::pivot_longer(!c(dt, date), 
                      names_to = c("Variable", "Plot", "Depth", "House"), 
                      names_pattern = "(.*)_P(.*)_(.*)_H(.*)") %>%
  mutate(House = factor(as.numeric(House)),
         Depth = factor(as.numeric(Depth)),
         Plot = factor(as.numeric(Plot)),
         ID = paste0(Variable, "_P", Plot, "_", Depth, "_H", House)) %>%
  arrange(ID, dt)

Ta <- read.csv("../data/Halfhourly_VWC_Weather_030621.csv") %>%
  dplyr::select(TIMESTAMP, T_inside, RH_inside, T_outside, RH_outside) %>%
  mutate(TIMESTAMP = as.POSIXct(TIMESTAMP),
         date = as.Date(TIMESTAMP),
         T_inside = ifelse(T_inside < -10, NA, T_inside),
         T_outside = ifelse(T_outside < -10, NA, T_outside),
         D_inside = RHtoVPD(RH_inside, T_inside),
         D_outside = RHtoVPD(RH_outside, T_outside)) %>%
  rename(dt = TIMESTAMP) %>%
  tidyr::pivot_longer(!c(dt, date), 
                      names_to = c("Variable", "Location"), 
                      names_pattern = "(.*)_(.*)") %>% 
  arrange(Variable, Location, dt)
```

Half-hourly data can be challenging to plot, so initial plots are created of daily mean/max/min.  

First, Ts at each of 28 house/plot combinations. Each house has 12 plots, of which 5 were instrumented in houses 1 and 5, 3 were instrumented in houses 2 and 4, and all 12 were instrumented in house 3. Of the instrumented plots, all 5 in houses 1 and 5 and 5 of 12 plots in house 3 were instrumented for all 3 depths. The remaining plots were only instrumented at the shallow (0-12 cm) depth. 
```{r, echo = FALSE}
# Soil temperature
Ts_daily <- Ts %>%
  group_by(ID, date) %>%
  summarise(Ts_mean = mean(value, na.rm = T),
            Ts_min = min(value, na.rm = T),
            Ts_max = max(value, na.rm = T),
            House = unique(House),
            Depth = unique(Depth),
            Plot = unique(Plot)) %>%
  mutate(plotID = paste0(House, "_", Plot))

cols <- brewer.pal(9, "BrBG")
# display.brewer.pal(9, "BrBG")
ggplot(Ts_daily, aes(x = date, y = Ts_mean, color = Depth)) +
  geom_errorbar(aes(ymin = Ts_min, ymax = Ts_max), width = 0, alpha = 0.2) +
  geom_point(size = 0.25) +
  scale_y_continuous(expression(paste(T[soil], " (", degree, "C)"))) +
  scale_x_date(date_labels = "%b") + 
  facet_wrap(~plotID, nrow = 7) + 
  theme_bw(base_size = 12) +
  scale_color_manual(labels = c("0-12 cm", "25 cm", "75 cm"), values = cols[1:3]) +
  theme(axis.title.x = element_blank(),
        strip.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
```

Next, soil water potential at two plots. Two plots within house 3 were instrumented for SWP at all three depths. It appears that these 2 plots were subjected to different treatments.
```{r, echo = FALSE}
# Soil water potential
WP_daily <- WC %>%
  filter(Variable == "SWP") %>%
  group_by(ID, date) %>%
  summarise(WP_mean = mean(value, na.rm = TRUE),
            WP_min = min(value, na.rm = TRUE),
            WP_max = max(value, na.rm = TRUE),
            House = unique(House),
            Depth = unique(Depth),
            Plot = unique(Plot)) %>%
  mutate(plotID = paste0(House, "_", Plot),
         WP_min = ifelse(!is.finite(WP_min), NA, WP_min),
         WP_max = ifelse(!is.finite(WP_max), NA, WP_max)) %>%
  ungroup()

cols <- brewer.pal(11, "Spectral")
# display.brewer.pal(11, "Spectral")
ggplot(WP_daily, aes(x = date, y = WP_mean/1000, color = Depth)) +
  geom_errorbar(aes(ymin = WP_min/1000, ymax = WP_max/1000), width = 0, alpha = 0.2) +
  geom_point(size = 0.25) +
  scale_y_continuous(expression(paste(Psi[soil], " (MPa)"))) +
  scale_x_date(date_labels = "%b") + 
  facet_wrap(~plotID, nrow = 2) + 
  theme_bw(base_size = 12) +
  scale_color_manual(labels = c("0-12 cm", "25 cm", "75 cm"), values = cols[11:9]) +
  theme(axis.title.x = element_blank(),
        strip.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
# ggplotly()
```

Next, soil water content at each of 28 house/plot combinations. These were instrumented to match the soil temperature probes. The temporal repackaging watering treatments are particularly apparent at the shallow depth. 
```{r, echo = FALSE}
# Soil water content
WC_daily <- WC %>%
  filter(Variable == "VWC") %>%
  group_by(ID, date) %>%
  summarise(WC_mean = mean(value, na.rm = TRUE),
            WC_min = min(value, na.rm = TRUE),
            WC_max = max(value, na.rm = TRUE),
            House = unique(House),
            Depth = unique(Depth),
            Plot = unique(Plot)) %>%
  mutate(plotID = paste0(House, "_", Plot),
         WC_min = ifelse(!is.finite(WC_min), NA, WC_min),
         WC_max = ifelse(!is.finite(WC_max), NA, WC_max)) %>%
  ungroup()

cols <- brewer.pal(11, "Spectral")
# display.brewer.pal(11, "Spectral")
ggplot(WC_daily, aes(x = date, y = WC_mean, color = Depth)) +
  geom_errorbar(aes(ymin = WC_min, ymax = WC_max), width = 0, alpha = 0.2) +
  geom_point(size = 0.25) +
  scale_y_continuous(expression(paste(Theta[soil], "( ", m^3, " ", m^-3, ")"))) +
  scale_x_date(date_labels = "%b") + 
  facet_wrap(~plotID, nrow = 7) + 
  theme_bw(base_size = 12) +
  scale_color_manual(labels = c("0-12 cm", "25 cm", "75 cm"), values = cols[11:9]) +
  theme(axis.title.x = element_blank(),
        strip.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
# ggplotly()
```

Finally, VPD, RH, and air temperature are plotted for inside and outside of the houses. The large degree of overlap confirms that the houses are likely to have minimal impact on micrometeorological conditions. 
```{r, echo = FALSE}
# Ta: Temp, RH, VPD
Ta_daily <- Ta %>%
  group_by(Variable, Location, date) %>%
  summarise(mean = mean(value, na.rm = TRUE),
            min = min(value, na.rm = TRUE),
            max = max(value, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(min = ifelse(!is.finite(min), NA, min),
         max = ifelse(!is.finite(max), NA, max),
         var = case_when(Variable == "T" ~ "T~(~degree~C)",
                         Variable == "RH" ~ "RH~('%')",
                         Variable == "D" ~ "D~(kPa)"),
         Location = factor(Location, levels = c("outside", "inside")))

ggplot(Ta_daily, aes(x = date, y = mean, color = Location)) +
  geom_errorbar(aes(ymin = min, ymax = max), width = 0, alpha = 0.1) +
  geom_point(size = 0.25, alpha = 0.75) +
  scale_x_date(date_labels = "%b") + 
  facet_wrap(~ var, nrow = 7, strip.position = "left", labeller = label_parsed,
             scales = "free_y") + 
  theme_bw(base_size = 12) +
  theme(axis.title.x = element_blank(),
        strip.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.title.y = element_blank(),
        strip.placement = "outside",)
# ggplotly()
```

## Matching soil water content and soil water potential

For the two plots with both kinds of sensors, we combine the daily data and plot as a time-series to obtain the time periods of correlation. Rectangles denoting the major treatment dates and transitions are added. 

```{r, echo = FALSE}
SW_long <- WC_daily %>%
  select(-ID) %>%
  filter(plotID %in% unique(WP_daily$plotID)) %>%
  left_join(y = select(WP_daily, -ID, -plotID), 
            by = c("date", "House", "Depth", "Plot")) %>%
  mutate(ID = paste0("P", Plot, "_", Depth, "_H", House)) %>%
  relocate(date, plotID, ID, Plot, Depth, House) %>%
  tidyr::drop_na() %>%
  tidyr::pivot_longer(!c(1:6), 
                      names_to = c("Variable", "Type"), 
                      names_pattern = "(.*)_(.*)") %>%
  mutate(value = ifelse(Variable == "WP", value/1000, value)) %>%
  tidyr::pivot_wider(names_from = "Type") %>%
  mutate(var = case_when(Variable == "WC" ~ "Theta~(m^3~m^-3)",
                         Variable == "WP" ~ "Psi~(MPa)"))

# Add dates
timing <- data.frame(Period = rep(c("Winter 2020", "Premonsoon", "S1-S4", "Winter 2021"), 4),
                     st = rep(c(as.Date("2019-11-01"), as.Date("2020-05-01"), 
                            as.Date("2020-07-13"), as.Date("2020-10-01")),4),
                     en = rep(c(as.Date("2020-04-30"), as.Date("2020-07-12"),
                            as.Date("2020-09-30"), as.Date("2021-03-06")),4),
                     var = rep(c("Theta~(m^3~m^-3)", "Psi~(MPa)"), each = 8),
                     plotID = rep(c("3_4", "3_7"), each = 4))
timing$Period <- factor(timing$Period, levels = c("Winter 2020", "Premonsoon", "S1-S4", "Winter 2021"))

cols <- brewer.pal(11, "Spectral")
# display.brewer.pal(11, "Spectral")
ggplot() +
  geom_rect(data = timing, 
            aes(xmin = st, xmax = en, ymin = -Inf, ymax = Inf, fill = Period), 
            alpha = 0.2) +
  geom_errorbar(data = SW_long, 
                aes(x = date, color = Depth, ymin = min, ymax = max), alpha = 0.5) +
  geom_point(data = SW_long,
             aes(x = date, y = mean, color = Depth), size = 0.5) +
  facet_grid(var~plotID, labeller = labeller(var = label_parsed),
             scales = "free", switch = "y") + 
  scale_color_manual(labels = c("0-12 cm", "25 cm", "75 cm"), values = cols[11:9]) +
  theme_bw(base_size = 12) +
  theme(axis.title.x = element_blank(),
        strip.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.title.y = element_blank(),
        strip.placement = "outside")
```

We plot the data 1:1 for each of the 4 periods. 

```{r, echo = FALSE}
tm <- timing[1:4, 1:3]
SW_wide <- WC_daily %>%
  select(-ID) %>%
  filter(plotID %in% unique(WP_daily$plotID)) %>%
  left_join(y = select(WP_daily, -ID, -plotID), 
            by = c("date", "House", "Depth", "Plot")) %>%
  mutate(ID = paste0("P", Plot, "_", Depth, "_H", House),
         WP_mean = WP_mean/1000,
         WP_min = WP_min/1000,
         WP_max = WP_max/1000,
         depth = case_when(Depth == 1 ~ "0-12 cm",
                           Depth == 2 ~ "25 cm",
                           Depth == 3 ~ "75 cm",)) %>%
  relocate(date, plotID, ID, Plot, Depth, House) %>%
  tidyr::drop_na() %>%
  mutate(Period = case_when(date <= tm$en[1] & date >= tm$st[1] ~ tm$Period[1],
                            date <= tm$en[2] & date >= tm$st[2] ~ tm$Period[2],
                            date <= tm$en[3] & date >= tm$st[3] ~ tm$Period[3],
                            date <= tm$en[4] & date >= tm$st[4] ~ tm$Period[4]))
cols <- brewer.pal(11, "Spectral")
# display.brewer.pal(11, "Spectral")
P4_left <- ggplot(filter(SW_wide, plotID == "3_4"),
                  aes(x = WP_mean, y = WC_mean, col = Depth)) +
  geom_errorbar(aes(ymin = WC_min, ymax = WC_max), alpha = 0.2) +
  geom_errorbarh(aes(xmin = WP_min, xmax = WP_max), alpha = 0.2) +
  geom_point(size = 0.25, alpha = 0.75) +
  scale_x_continuous(expression(paste(Psi, " (MPa)"))) +
  scale_y_continuous(expression(paste(Theta, " (", m^3, " ", m^-3, ")"))) +
  facet_wrap(~Period, scales = "free", ncol = 1) + 
  scale_color_manual(labels = c("0-12 cm", "25 cm", "75 cm"), values = cols[11:9]) +
  theme_bw(base_size = 12) +
  theme(strip.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = c(0.25, 0.95),
        legend.background = element_blank(),
        legend.title = element_blank())

P4_right <- ggplot(filter(SW_wide, plotID == "3_4"), 
                   aes(x = date, color = Depth)) +
  geom_pointrange(aes(y = WP_mean, ymin = WP_min, ymax = WP_max, shape = "WP"), size = 0.25) +
  geom_pointrange(aes(y = WC_mean*10, ymin = WC_min*10, ymax = WC_max*10, shape = "VWC*10"), 
                  size = 0.25) +
  scale_y_continuous(expression(paste(Psi, " (MPa) | ", Theta, "*10 (", m^3, " ", m^-3, ")"))) +
  facet_wrap(~Period, scales = "free", ncol = 1) + 
  scale_color_manual(labels = c("0-12 cm", "25 cm", "75 cm"), values = cols[11:9]) +
  theme_bw(base_size = 12) +
  theme(axis.title.x = element_blank(),
        strip.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  guides(color = "none",
         shape = "none")
P7_left <- ggplot(filter(SW_wide, plotID == "3_7"),
                  aes(x = WP_mean, y = WC_mean, col = Depth)) +
  geom_errorbar(aes(ymin = WC_min, ymax = WC_max), alpha = 0.2) +
  geom_errorbarh(aes(xmin = WP_min, xmax = WP_max), alpha = 0.2) +
  geom_point(size = 0.25, alpha = 0.75) +
  scale_x_continuous(expression(paste(Psi, " (MPa)"))) +
  scale_y_continuous(expression(paste(Theta, " (", m^3, " ", m^-3, ")"))) +
  facet_wrap(~Period, scales = "free", ncol = 1) + 
  scale_color_manual(labels = c("0-12 cm", "25 cm", "75 cm"), values = cols[11:9]) +
  theme_bw(base_size = 12) +
  theme(strip.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = c(0.25, 0.95),
        legend.background = element_blank(),
        legend.title = element_blank())

P7_right <- ggplot(filter(SW_wide, plotID == "3_7"), 
                   aes(x = date, color = Depth)) +
  geom_pointrange(aes(y = WP_mean, ymin = WP_min, ymax = WP_max, shape = "WP"), size = 0.25) +
  geom_pointrange(aes(y = WC_mean*10, ymin = WC_min*10, ymax = WC_max*10, shape = "VWC*10"), 
                  size = 0.25) +
  scale_y_continuous(expression(paste(Psi, " (MPa) | ", Theta, "*10 (", m^3, " ", m^-3, ")"))) +
  facet_wrap(~Period, scales = "free", ncol = 1) + 
  scale_color_manual(labels = c("0-12 cm", "25 cm", "75 cm"), values = cols[11:9]) +
  theme_bw(base_size = 12) +
  theme(axis.title.x = element_blank(),
        strip.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  guides(color = "none",
         shape = "none")

jpeg("../plots/3_4_byseason_daily.jpg", width = 6, height = 8, 
     units = "in", res = 600)
plot_grid(P4_left, P4_right, ncol = 2)
dev.off()

jpeg("../plots/3_7_byseason_daily.jpg", width = 6, height = 8, 
     units = "in", res = 600)
plot_grid(P7_left, P7_right, ncol = 2)
dev.off()
```

Save out daily tables as .Rdata files for future use in Shiny app. 
```{r}
save(Ts_daily, file = "../soildata_app/soilTempDaily.Rdata")
save(WC_daily, file = "../soildata_app/soilWaterContentDaily.Rdata")
save(WP_daily, file = "../soildata_app/soilWaterPotentialDaily.Rdata")
save(Ta_daily, file = "../data/weatherVariablesDaily.Rdata")
save(SW_long, file = "../soildata_app/soilWC-WPDaily.Rdata")
```