---
title: "Model report"
author: "Jessica Guo"
date: "4/21/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(udunits2)
library(dplyr)
library(coda)
library(broom.mixed)
library(ggplot2)
```


#### Introduction
The temporal repackaging of rainfall experiment at the Santa Rita Experimental Range ('RainManSR') utilizes a random block design to test the influence of rainfall intensity and frequency on soil, plant, and ecosystem dyanamics, while holding total rainfall amount constant. The soil water content (SWC) of most plots are monitored at three depths 0-12 cm, 25 cm, and 75 cm. In two individual plots representing the S1 (frequent small rain events) and the S4 (infrequent large rain events), soil water potential (SWP) probes were additionally installed, allowing for us to generate <i>in-situ</i> soil moisture release curves. 


#### Data description
All soil moisture data can be interactively explored via the [Shiny app](https://viz.datascience.arizona.edu/rainman-soildata/). For this analysis, we selected the 25 cm depth, for which A) a range of moisture values was achieved and B) the SWC and SWP sensors were installed at identical depths. For both the S1 and S4 treatments, we used the late winter and early spring data (2019-12-01 to 2020-04-30) for the 25 cm depth (n = 152 for each S1 and S4). 


#### Model description

We used the Van Genuchten function to predict SWC from SWP. The observed SWC($\Theta_i$) were normally distributed ($i = 1 \dots 304$):

$\Theta_i \sim Normal(\mu_i, \tau)$

The expected value was modeled non-linearly as:

$\mu_i = \Theta_{r, trt(i)} + \frac{\Theta_s - \Theta_{r, trt(i)}}{[1 + (\alpha_{trt(i)}\Psi_i)^{n_{trt(i)}}]^{1-\frac{1}{n_{trt(i)}}}}$

Where $\Theta_s$ was a constant 0.44 (porosity of loamy sand), and the parameters $\Theta_r$, $\alpha$, and $n$ were indexed by $trt(i)$ and allowed to vary by the irrigation treatment.

Semi-informative but wide priors were specified:

$\Theta_{r, t} ~ \sim Uniform(0.01, 0.05)$

$\alpha_t \sim Uniform(0, 50)$

$n_t ~ \sim Uniform(0.5, 3)$

The above equations were programmed in JAGS; three parallel chain were run for 150,000 iterations with thinning by 50 to reduce storage space for a total of 9000 samples. Parameters are summarized as posterior means and the central 95% credible interval (CI). If the posterior mean of one parameter is not overlapped by the CI of the other and vice versa, then the two parameters are said to be significantly different. 

```{r, echo=FALSE}
# Load data 
load("out.Rdata")

dat <- filter(out, Depth == 2) %>%
  mutate(pressure_head_cm = ud.convert(WP_mean, "MPa", "cm_H2O"),
         pressure_head_m = ud.convert(pressure_head_cm, "cm", "m"))

# Load codas
load(file = "coda/jm_coda.Rdata")
load(file = "coda/jm_rep.Rdata")

# Tidy parameters
sum_param <- tidyMCMC(as.mcmc(jm_coda[[1]], jm_coda[[2]], jm_coda[[3]]),
                      conf.int = TRUE,
                      conf.method = "quantile")

# # Tidy replicated and add to original data
sum_rep <- tidyMCMC(as.mcmc(jm_rep[[1]], jm_rep[[2]], jm_rep[[3]]),
                    conf.int = TRUE,
                    conf.method = "quantile")

pred <- cbind.data.frame(dat, sum_rep)
```


#### Model fit

The observed vs. predicted indicate good model fit for both treatments, R<sup>2</sup> = 0.98. 

```{r, echo=FALSE, fig.height=4, fig.width=8}
m1 <- summary(lm(WC_mean ~ estimate, data = filter(pred, Summer == "S1")))
m4 <- summary(lm(WC_mean ~ estimate, data = filter(pred, Summer == "S4")))

# Observed vs. fitted
pred %>%
  ggplot(aes(x = WC_mean, y = estimate)) +
  geom_abline(slope = 1, intercept = 0, lty = 2) +
  geom_pointrange(aes(ymin = conf.low, 
                      ymax = conf.high,
                      col = Summer)) +
  scale_color_manual(values = c("goldenrod", "forestgreen")) +
  scale_x_continuous("Observed SWC") +
  scale_y_continuous("Predicted SWC") +
  facet_wrap(~Summer) +
  theme_bw(base_size = 14) +
  theme(legend.title = element_blank(),
        legend.position = c(0.9, 0.2))

```
 
Model fit also looks promising when plotted along the original Van Genuchten axes of water content vs. water potential. 

```{r, echo=FALSE, fig.height=4, fig.width=8}
pred %>% 
  ggplot(aes(x = -pressure_head_m)) +
  geom_point(aes(y = WC_mean, 
                 color = "Observed")) +
  geom_pointrange(aes(y = estimate, 
                      ymin = conf.low,
                      ymax = conf.high,
                      color = "Predicted")) +
  facet_wrap(~Summer) +
  scale_color_manual(values = c("black", "gray")) +
  scale_x_log10(name = "Pressure Head (-m)",
              breaks = c(0.01, 0.1, 1, 10, 100, 1000),
              labels = c(0.01, 0.1, 1, 10, 100, 1000)) +
  scale_y_continuous("SWC") +
  theme_bw(base_size = 14) +
  theme(legend.title = element_blank(),
        legend.position = c(0.85, 0.85))
```

Parameters $\Theta_r$, $\alpha$, and $n$ were estimated by the model and were all found to be significantly different between the two treatments, S1 and S4. 

```{r, echo=FALSE, fig.height=3, fig.width=8}
sum_param %>%
  mutate(parameter = sub("\\[[0-9]\\]", "", term),
         Treatment = case_when(grepl("1", term) ~ "S1",
                            grepl("2", term) ~ "S4"),
         parameter = factor(parameter, levels = c("theta.r", "alpha.cm", "n"))) %>%
  filter(parameter %in% c("alpha.cm", "n", "theta.r")) %>%
  ggplot(aes(x = Treatment, y = estimate)) +
  geom_pointrange(aes(ymin = conf.low, 
                      ymax = conf.high)) +
  scale_y_continuous("Posterior estimate") +
  facet_wrap(~parameter, scales = "free") +
  theme_bw(base_size = 14)
```

If we adopted the posterior mean of each parameter, our data set could only validate the dry end of the Van Genuchten relationship. Nevertheless, the parameters seem appropriate. 

```{r, echo=FALSE, fig.height=4, fig.width=8}
param1 <- sum_param %>%
  filter(grepl("\\[1\\]", term))

param4 <- sum_param %>%
  filter(grepl("\\[2\\]", term))

VG <- function(h.cm, vec){
  alpha <- as.numeric(vec[1])
  n <- as.numeric(vec[2])
  theta.r <- as.numeric(vec[3])

  wc <- theta.r + ((0.44 - theta.r)/
                            (1 + (alpha*h.cm)^n)^(1-(1/n)))
  
  return(wc)
}

out1 <- data.frame(ph_cm = seq(1, 10000, by = 1),
                   WC = VG(seq(1, 10000, by = 1), 
                           vec = param1$estimate),
                   Summer = "S1")

out4 <- data.frame(ph_cm = seq(1, 10000, by = 1),
                   WC = VG(seq(1, 10000, by = 1), 
                           vec = param4$estimate),
                   Summer = "S4")

ggplot() +
  geom_line(data = out1,
            aes(x = ph_cm, y = WC, color = Summer),
            lwd = 1) +
  geom_line(data = out4,
            aes(x = ph_cm, y = WC, color = Summer),
            lwd = 1) +
  geom_point(data = dat, 
             aes(x = -pressure_head_cm,
                 y = WC_mean),
             alpha = 0.5) +
  scale_x_log10(name = "Pressure Head (-cm)",
                breaks = c(0.01, 0.1, 1, 10, 100, 1000, 10000),
                labels = c(0.01, 0.1, 1, 10, 100, 1000, 10000)) +
  scale_y_continuous("SWC") +
  scale_color_manual(values = c("goldenrod", "forestgreen")) +
  facet_wrap(~Summer) +
  theme_bw(base_size = 14) +
  theme(legend.title = element_blank(),
        legend.position = c(0.9, 0.8))

```
